Using functor categories to generate intermediate code with Agda

### 2025-05-19 draft.tex
Minitoc is back! Hooray!

\documentclass[12pt,a4paper]{report}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathabx}
\usepackage{stmaryrd}
\usepackage{upgreek}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{enumerate}
\usepackage[vmargin=2.5cm, hmargin=2.5cm]{geometry}
\usepackage{hyperref}
\usepackage{fontspec}
\usepackage{unicode-math}
\usepackage{stackengine}
\usepackage{minitoc}
\usepackage{amsthm}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{xeCJK}
\usepackage{dirtree}
\usepackage{float}
\usepackage{pdfpages}
\usepackage{atbegshi}
\usepackage{multirow, makecell}
\usepackage{enumitem}
\usepackage{hhline}

@@

@@

\begin{titlepage}
    \vspace*{5em}
    \textbf{\LARGE Declaration of Originality}
    \vspace{2em}

@@

\begin{titlepage}
    \vspace*{5em}
    \textbf{\LARGE Proforma}
    \vspace{2em}

    \begin{tabular}{ll}
        \textbf{Candidate Number:} & 2330G \\
        \textbf{Title of Project:} & Using functor categories to generate intermediate code with Agda \\
        \textbf{Examination} & Computer Science Tripos - Part II - 2025 \\
        \textbf{Word-count:} & [wordcount] \footnotemark \\
        \textbf{Code line count:} & [linecount] \footnotemark \\
        \textbf{Project Originator:} & Yulong Huang \\
        \textbf{Project Supervisor:} & Yulong Huang and Yufeng Li \\
    \end{tabular}

@@

\mainmatter
\chapter{Introduction}
    \minitoc
    Programming languages act as a bridge between human thoughts and machine execution. They exist in two complementary realms: the human realm, where intent is expressed through abstractions like variables and functions, and the machine realm, where low-level instructions are executed on hardware.

    Denotational semantics is related to the human realms. It provides a theoretical framework for defining the meaning of programming languages by interpreting them into mathematical objects. By formalising the denotational semantics of a programming language, we ensure that its abstraction align with our intentions.

    Compilers are related to the machine realms. They are programs that translate high-level code into executable instructions, resolving abstractions into concrete operations like memory allocation and register management. Compilers are essential for turning human-written code into physical computation.

@@

        \motivation{Implementation of the compiler}
        Reynolds concluded that he did not have a proper dependently typed programming language in hand, so his compiler remained a partial function~\autocite[Ch.6]{Reynolds}. I aim to provide a computer implementation of this theoretical framework in a dependently typed programming language.

    \section{Language choice: Agda's advantages}
        Agda~\autocite{Agda} is a dependently typed programming language and proof assistant. As a dependently typed language, Agda captures the source language's intrinsic syntax with indexed families, which contains only well-typed terms. Therefore, it focuses on the correct programs and rules out the ill-typed nonsensical inputs.

@@

        The project successfully met all of the above success criteria, addressed the two motivations presented in~\secref{sec: motivation} and contributed to the following:
        \begin{quote}            
            \savedmotivationI
            I formalised the terms in the source language and target language in Agda. I also refined some type definitions in the denotational semantics of the source language.

            \savedmotivationII
            I implemented the compiler from the source language to the target language in Agda. 
        \end{quote}

\chapter{Preparation}
    \minitoc

    For implementing the compiler, I need to understand the theoretical background of presheaves and functor categories that are used as the denotational semantics of the source language, and I need to understand dependent types to correctly express the dependent function space of presheaf exponentials.

    This chapter begins with my starting point and introducing core concepts of category theory. It then provides a brief overview of Agda as a dependently typed programming language, concluding with a requirements analysis for the compiler design.

@@

            \begin{definition}[Exponential]
                Given a category $\mathcal{C}$ with binary products, the \emph{exponential} of two objects $X$ and $Y$ in $\mathcal{C}$ is specified by
                \begin{itemize}
                    \item a $\mathcal{C}$-object $X \Rightarrow Y$;
                    \item a $\mathcal{C}$-morphism $\textbf{app} : (X \Rightarrow Y) \times X \to Y$ called the \emph{application} of $X \Rightarrow Y$;
                \end{itemize}
                such that for all $Z \in \textbf{obj}(\mathcal{C})$ and morphisms $f : Z \times X \to Y$, there exists a unique morphism $u : Z \to X \Rightarrow Y$ such that the following diagram commutes in $\mathcal{C}$:
                \begin{equation}
                    \vcenter{\hbox{\includegraphics[width=0.4\textwidth]{exponential.pdf}}}
                    \label{cd: exponential}
                \end{equation}
                We write $\textbf{cur} f$ for the unique morphism $u$ such that $f = \textbf{app} \circ (\textbf{cur} f \times \textbf{id}_X)$, 
                where $\textbf{cur} f$ is called the \emph{currying} of $f$.
            \end{definition}
            It can be shown that the exponential is unique up to (unique) isomorphism.

@@

@@

        \begin{table}[H]
            \centering
            \begin{tabular}{|l|l|l|}
                \hline
                \textbf{File} & \textbf{Description}\\
                \hhline{|=|=|}
                \texttt{lib.agda} & shared utilities (e.g. natural numbers and operations, equality, etc.) \\
                \hline
                \texttt{source.agda} & source language syntax \\
                \hline
                \texttt{target.agda} & target language (stack machine) instructions \\
                \hline
                \texttt{compiler.agda} & compiler derived from denotational semantics \\
                \hline
                \texttt{test.agda} & test cases for the compiler \\
                \hline
            \end{tabular}
            \caption{File descriptions}
            \label{tab: file_descriptions}
        \end{table}

@@

    \subsubsection{Properties}
    The properties of stack descriptors can be summarised as follows:
    \begin{table}[H]
        \centering
        \begin{tabular}{|c|c|}
            \hline
            \textbf{Agda term} & \textbf{Mathematical meaning} \\
            \hhline{|=|=|}
            \mbt{≤ₛ-refl} & $\forall sd.\ sd \leq_\mathsf{s} sd$ \\
            \hline
            \mbt{≤ₛ-trans} & $\forall sd, sd', sd''.\ sd \leq_\mathsf{s} sd' \land sd' \leq_\mathsf{s} sd'' \Rightarrow sd \leq_\mathsf{s} sd''$ \\
            \hline
            \mbt{+ₛ→≤ₛ} & $\forall sd, n.\ sd \leq_\mathsf{s} sd +_\mathsf{s} n$ \\
            \hline
            \mbt{sub-sd≤ₛ} & $\forall sd, sd', sd''. sd' \equiv sd'' \land sd \leq_\mathsf{s} sd' \Rightarrow sd \leq_\mathsf{s} sd''$ \\
            \hline
            \mbt{–ₛ≡} & $\forall f, d, d′, n.\ d′ - n \equiv d \Rightarrow \ang{f, d} \equiv \ang{f, d′} -_\mathsf{s} n$ \\
            \hline
        \end{tabular}
        \caption{Properties of order and addition of stack descriptors}
        \label{tab: stack_descriptor_properties}
    \end{table}
    % We can use reflexivity and transitivity of the order $<$ and $\leq$ to prove that the order of stack descriptors $\leq_\mathsf{s}$ is a preorder. The proof in Agda is as follows:
    % \begin{proof}
    %     We need to show that $\leq_\mathsf{s}$ is reflexive and transitive. 
    %     \begin{itemize}
    %         \item 
    %             Reflexivity: $\textit{sd} \leq_\mathsf{s} \textit{sd}$ is trivially true since $f = f$ and $d \leq d$.
    %         \item
    %             Transitivity: if $\ang{f, d} \leq_\mathsf{s} \ang{f', d'}$ and $\ang{f', d'} \leq_\mathsf{s} \ang{f'', d''}$, we do the following case split:
    %             \begin{enumerate}
    %                 \item 
    %                     If $f < f'$, $f' = f''$ and $d \leq d'$, then $f < f''$; 
    %                 \item 
    %                     If $f < f'$ and $f' < f''$, then $f < f''$ by transitivity of $<$;
    %                 \item
    %                     If $f = f'$, $d \leq d'$ and $f' < f''$, then $f < f''$;
    %                 \item
    %                     Otherwise $f = f''$ and $d \leq d''$ by transitivity of $\leq$. \footnote{Note that the properties of $=$ is not required in as in the Agda definition of $\leq_\mathsf{s}$, we directly define $f'$ to be $f$ instead of defining a equivalence relation, thus equivalence can be checked directly via type checking.}
    %             \end{enumerate}
    %     \end{itemize}
    % \end{proof}

    % The proof is defined in Agda as follows:
    % \[\includegraphics[scale=1.2]{/target/stack_descriptor_order_proof.pdf}\]

    % When a natural number is added to a stack descriptor, it is guaranteed that the new stack descriptor is not less than the old stack descriptor. This can be proved directly using the same properties of integer addition. The proof is defined in Agda as follows:
    % \[\includegraphics[scale=1.2]{/target/stack_descriptor_add_proof.pdf}\]

    \subsection{Grammar} \label{subsec: grammar}
    The grammar of the target language\footnote{This grammar can be extended with subroutines as an \textsf{L} command, and conditionals as an \textsf{I} command} is defined as follows, given the current stack descriptor $\textit{sd} = \ang{f, d}$, and new stack descriptor $\textit{sd′}$ and $\textit{sd\textsuperscript{v}}$:
    \begin{figure}[H]
        \centering
        \[\begin{aligned}
            \bracket{\textsf{L}_{\textit{sd}}} ::={}& \textit{sd\textsuperscript{v}} \quad \text{when } \textit{sd\textsuperscript{v}} \leq_\mathsf{s} \textit{sd} \\
                            % &\mid \texttt{sbrs}  \\
            \bracket{\textsf{S}_{\textit{sd}}} ::={}& \bracket{\textsf{L}_{\textit{sd}}} \\
                            &\mid \texttt{lit } \bracket{\text{integer}} \\
            \bracket{\textsf{R}_{\textit{sd}}} ::={}& \bracket{\textsf{S}_{\textit{sd}}} \\
                            &\mid \bracket{ \text{unary operator} } \bracket{ \textsf{S}_{\textit{sd}}}\\
                            &\mid \bracket{\textsf{S}_{\textit{sd}}} \bracket{\textsf{binary operator}} \bracket{ \textsf{S}_{\textit{sd}} } \\
            \bracket{\textsf{I}_{\textit{sd}}} ::={}& \texttt{stop} \\
                            &\left.
                            \begin{aligned}
                            &|{\ } \bracket{\textsf{L}_{\textit{sd}+\delta}} := \bracket{\textsf{R}_{\textit{sd}}}[\delta] \text{ ; } \bracket{\textsf{I}_{\textit{sd}+\delta}} \\
                            % &|{\ } \texttt{if } \bracket{\textsf{S}_{\textit{sd}}} \bracket{\text{relational operator}} \bracket{\textsf{S}_{\textit{sd}}}[\delta] \\
                            % & \hphantom{\mid {}} \texttt{ then } \bracket{\textsf{I}_{\textit{sd}+\delta}} \texttt{ else } \bracket{\textsf{I}_{\textit{sd}+\delta}} \\
                            &|{\ } \texttt{adjustdisp} [\delta] \text{ ; } \bracket{\textsf{I}_{\textit{sd}+\delta}} \\
                            \end{aligned}
                            \quad \right\} \text{if } d + \delta \geq 0  \\
                            &{}\mid{} \texttt{popto } \textit{sd′} \text{ ; } \bracket{\textsf{I}_{\textit{sd′}}} \quad \text{when } \textit{sd′} \leq_\mathsf{s} \textit{sd} \\
        \end{aligned}\]
        \caption{Target language grammar}
        \label{fig: target_grammar}
    \end{figure}
    Some explanations of the grammar are as follows:
    \begin{itemize}
        % \item 
        %     $\texttt{sbrs}$: a register used to communicate the result of function procedures implemented by closed subroutines.
        \item 
            $\texttt{lit}$: a literal integer.
        \item 
            $\texttt{stop}$: stops the execution of the program.
        \item 
            $\delta$: an integer representing the displacement adjustment.
        \item 
            $\texttt{adjustdisp}$: adjusts the displacement of specified stack descriptor.
        \item 
            $\texttt{popto}$: pops to the specified stack descriptor, which is used to reduce the number of frames.
    \end{itemize}
    Note that the design of right-hand sides ensures that the right operand of the assignment contains at most one operator, which means that any instruction that involves multiple operators must be divided into multiple instructions. This feature of the target language significantly complicates the compilation of integer expressions, which we will analysis in detail in~\secref{subsubsec: compiler_intexp}.

@@

    \section{Customised Library} \label{sec: lib}
    The customised library is a collection of types and functions centred around natural numbers, their operations and properties. Instead of using the standard library~\autocite{agda_std}, I implemented a customised library for the following reasons:
    \begin{itemize}
        \item
            \textbf{Specialised Requirements}: The standard library's definitions (e.g. subtraction as monus, `$\dotdiv$') do not align with my project's need for a rigorously defined subtraction operation. This is specified in~\secref{subsec: subtraction}.
        \item
            \textbf{Verification Clarity}: By defining the basic data types and functions from scratch, I can tailor properties (e.g. the property `$n-[n-m] \equiv m$') to my specific requirements, and avoid unnecessary complexity (e.g. `$\leq$' being also defined as a boolean predicate in the standard library). 
        \item
            \textbf{Minimal Dependencies}: Avoiding the standard library reduces external assumptions, making the project self-contained. 
    \end{itemize}
    All definitions and properties can be found in Appendix~[TBD].

    \subsection{Definitions from standard library}
    The following is a list of definitions (or their variants) from the standard library and PLFA~\cite{plfa} that are redefined in the customised library:
    \begin{itemize}
        \item 
            Natural numbers ($\mb{\bN}$), integers ($\mb{\bZ}$) and addition (\mbt{+}).
        \item
            Equality (\mbt{≡}), congruence ($\mb{\mathsf{cong}}$), substitution ($\mb{\mathsf{sub}}$) and transitivity of equality ($\mb{\mathsf{trans}}$).
        \item
            Order of natural numbers, including \mbt{≤} and \mbt{<}.
        \item
            Properties of preorder and addition (summarised in Table~\ref{tab: properties}).
    \end{itemize}
    \begin{table}[H]
        \centering
        \begin{tabular}{|l|l|}
            \hline
            \textbf{Agda term} & \textbf{Mathematical meaning} \\
            \hhline{|=|=|}
            \mbt{+-identity\textsuperscript{r}} & $\forall n \in \bN.\ n + 0 = n$ \\
            \hline
            \mbt{+-suc\textsuperscript{r}}& $\forall m, n \in \bN.\ m + \textsf{suc } n = \textsf{suc } (m + n)$ \\
            \hline
            \mbt{+-comm\textsuperscript{r}} & $\forall m, n \in \bN.\ m + n \equiv n + m$ \\
            \hline
            \mbt{≤-irrelevant} & For all $m, n \in \bN$, we consider all proofs of $m \leq n$ to be equal. \\
            \hline
            \mbt{≤-refl} & $\forall n \in \bN.\ n \leq n$ \\
            \hline
            \mbt{≤-trans} & $\forall m, n, p \in \bN.\ m \leq n \land n \leq p \Rightarrow m \leq p$ \\
            \hline
            \mbt{n≤suc-n} & $\forall n \in \bN.\ n \leq \textsf{suc } n$ \\
            \hline
            \mbt{m≡n,p≤n}\mb{$\to$}\mbt{p≤m} & $\forall p, m, n \in \bN.\ m \equiv n \land p \leq n \Rightarrow p \leq m$ \\
            \hline
            \mbt{+}\mb{$\to$}\mbt{≤} & $\forall m, n \in \bN.\ m \leq m + n$ \\
            \hline
            \mbt{+}\mb{$\to$}\mb{≤\textsuperscript{r}} & $\forall m, n \in \bN.\ m \leq n + m$ \\
            \hline
            \mbt{<-trans} & $\forall m, n, p \in \bN.\ m < n \land n < p \Rightarrow m < p$ \\
            \hline
            \mbt{<}\mb{$\to$}\mbt{s≤} & $\forall m, n \in \bN.\ m < n \Rightarrow \textsf{suc } m \leq n$ \\
            \hline
            \mbt{<}\mb{$\to$}\mbt{≤} & $\forall m, n \in \bN.\ m < n \Rightarrow m \leq n$ \\
            \hline
            \mbt{≤-compare} & $\forall m, n \in \bN.\ m \leq n \lor n \leq m$ \\
            \hline
        \end{tabular}
        \caption{Properties of order and addition}
        \label{tab: properties}
    \end{table}

    % \subsection{Natural numbers, integers and addition} \label{subsec: natural_numbers}
    % The definition of natural numbers, integers and addition is standard, and it is defined as follows:
    % \[\includegraphics[scale=1.2]{/lib/natural.pdf}\]
    % The use of \og{\textsf{BUILTIN}} is specified in~\secref{subsubsec: builtin_pragma}.

@@

        Please see Appendix~[TBD] for the full implementation of the function \textsf{NewVar}.

>>>>

\documentclass[12pt,a4paper]{report}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathabx}
\usepackage{stmaryrd}
\usepackage{upgreek}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{enumerate}
\usepackage[tmargin=2.3cm, hmargin=2.5cm, bmargin=2.3cm]{geometry}
\usepackage{hyperref}
\usepackage{fontspec}
\usepackage{unicode-math}
\usepackage{stackengine}
\usepackage{minitoc}
\usepackage{amsthm}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{xeCJK}
\usepackage{dirtree}
\usepackage{float}
\usepackage{pdfpages}
\usepackage{atbegshi}
\usepackage{multirow, makecell}
\usepackage{enumitem}
\usepackage{hhline}
\usepackage{titlesec}

@@

\titleformat{\chapter}[hang]
  {\normalfont\huge\bfseries}
  {\thechapter\quad }{0pt}{\huge}

\titlespacing*{\chapter}{0pt}{-25pt}{15pt}

\titlespacing*{\section}{0pt}{16pt}{8pt}

\titlespacing*{\subsection}{0pt}{12pt}{6pt}

\titlespacing*{\subsubsection}{0pt}{8pt}{4pt}

@@

\begin{titlepage}
    \textbf{\Large Declaration of Originality}
    \vspace{2em}

@@

\begin{titlepage}
    \textbf{\Huge Proforma}
    \vspace{2em}

    \begin{tabular}{ll}
        \textbf{Candidate Number:} & 2330G \\
        \textbf{Title of Project:} & Using functor categories to generate intermediate code with Agda \\
        \textbf{Examination} & Computer Science Tripos - Part II - 2025 \\
        \textbf{Word count:} & 9567 \footnotemark \\
        \textbf{Code line count:} & 1727 \footnotemark \\
        \textbf{Project Originator:} & Yulong Huang \\
        \textbf{Project Supervisor:} & Yulong Huang and Yufeng Li \\
    \end{tabular}

@@

\mainmatter
\chapter{Introduction}
    Programming languages act as a bridge between human thoughts and machine execution. They exist in two complementary realms: the human realm, where intent is expressed through abstractions like variables and functions, and the machine realm, where low-level instructions are executed on hardware.

    Denotational semantics is related to the human realm. It provides a theoretical framework for defining the meaning of programming languages by interpreting them into mathematical objects. By formalising the denotational semantics of a programming language, we ensure that its abstraction align with our intentions.

    Compilers are related to the machine realm. They are programs that translate high-level code into executable instructions, resolving abstractions into concrete operations like memory allocation and register management. Compilers are essential for turning human-written code into physical computation.

@@

        \motivation{Implementation of the compiler}
        Reynolds concluded that he did not have a proper dependently typed programming language in hand, so his compiler remained a partial function~\autocite[Ch.~6]{Reynolds}. I aim to provide a computer implementation of this theoretical framework in a dependently typed programming language.

    \section{Language choice: Agda's advantages}
        Agda~\autocite{Agda}, as a dependently typed proof assistant, captures source language's intrinsic syntax with indexed families. This automatically rules out the ill-typed nonsensical inputs, and only well-typed programs can be compiled.

@@

        The project successfully met all of the above success criteria and contributed to the following:
        \begin{quote}            
            \savedmotivationI
            I formalised the terms in the source and target language in Agda. I also refined some type definitions in the denotational semantics of the source language.
            \vspace*{1em}

            \savedmotivationII
            I implemented a compiler from the source language to the target language in Agda. 
        \end{quote}

\chapter{Preparation}
    \minitoc
    For implementing the compiler, I need to understand the theoretical background of presheaves and functor categories that are used as the denotational semantics of the source language, and I need to understand dependent types to correctly express the dependent function space of presheaf exponentials.

    This chapter begins with my starting point and introduces core concepts of category theory which form theoretical foundation of the compiler. It then provides a brief overview of Agda as a dependently typed programming language, concluding with a requirements analysis for the compiler design.

@@

            \begin{definition}[Exponential]
                Given a category $\mathcal{C}$ with binary products, the \emph{exponential} of two objects $X$ and $Y$ in $\mathcal{C}$ is specified by
                \begin{itemize}
                    \item a $\mathcal{C}$-object $X \Rightarrow Y$;
                    \item a $\mathcal{C}$-morphism $\textbf{app} : (X \Rightarrow Y) \times X \to Y$ called the \emph{application} of $X \Rightarrow Y$;
                \end{itemize}
                such that for all $Z \in \textbf{obj}(\mathcal{C})$ and morphisms $f : Z \times X \to Y$, there exists a unique morphism $u : Z \to X \Rightarrow Y$ such that the following diagram commutes in $\mathcal{C}$:
                \begin{equation}
                    \vcenter{\hbox{\includegraphics[width=0.3\textwidth]{exponential.pdf}}}
                    \label{cd: exponential}
                \end{equation}
                We write $\textbf{cur} f$ for the unique morphism $u$ such that $f = \textbf{app} \circ (\textbf{cur} f \times \textbf{id}_X)$, 
                where $\textbf{cur} f$ is called the \emph{currying} of $f$.
            \end{definition}
            It can be shown that the exponential is unique up to (unique) isomorphism.

@@

    % \clearpage

@@

        Each file in the repository is described in Table~\ref{tab: file_descriptions}. 
        \begin{table}[H]
            \centering
            \begin{tabular}{|l|l|l|}
                \hline
                \textbf{File} & \textbf{Description}\\
                \hhline{|=|=|}
                \texttt{lib.agda} & shared utilities (e.g. natural numbers and operations, equality, etc.) \\
                \hline
                \texttt{source.agda} & source language syntax \\
                \hline
                \texttt{target.agda} & target language (stack machine) instructions \\
                \hline
                \texttt{compiler.agda} & compiler derived from denotational semantics \\
                \hline
                \texttt{test.agda} & test cases for the compiler \\
                \hline
            \end{tabular}
            \caption{File descriptions}
            \label{tab: file_descriptions}
        \end{table}

@@

    \subsubsection{Properties}
    The properties of stack descriptors can be summarised as follows:
    \begin{table}[H]
        \centering
        \begin{tabular}{|c|c|}
            \hline
            \textbf{Agda term} & \textbf{Mathematical meaning} \\
            \hhline{|=|=|}
            \mbt{≤ₛ-refl} & $\forall sd.\ sd \leq_\mathsf{s} sd$ \\
            \hline
            \mbt{≤ₛ-trans} & $\forall sd, sd', sd''.\ sd \leq_\mathsf{s} sd' \land sd' \leq_\mathsf{s} sd'' \Rightarrow sd \leq_\mathsf{s} sd''$ \\
            \hline
            \mbt{+ₛ→≤ₛ} & $\forall sd, n.\ sd \leq_\mathsf{s} sd +_\mathsf{s} n$ \\
            \hline
            \mbt{sub-sd≤ₛ} & $\forall sd, sd', sd''. sd' \equiv sd'' \land sd \leq_\mathsf{s} sd' \Rightarrow sd \leq_\mathsf{s} sd''$ \\
            \hline
            \mbt{–ₛ≡} & $\forall f, d, d′, n.\ d′ - n \equiv d \Rightarrow \ang{f, d} \equiv \ang{f, d′} -_\mathsf{s} n$ \\
            \hline
        \end{tabular}
        \caption{Properties of order and addition of stack descriptors}
        \label{tab: stack_descriptor_properties}
    \end{table}
    The definitions of the above properties can be found in Appendix~\ref{app: target_properties}.

    \subsection{Grammar} \label{subsec: grammar}
    The grammar of the target language\footnote{This grammar can be extended with subroutines as an \textsf{L} command, and conditionals as an \textsf{I} command} is defined as follows, given the current stack descriptor $\textit{sd} = \ang{f, d}$, and new stack descriptor $\textit{sd′}$ and $\textit{sd\textsuperscript{v}}$:
    \begin{figure}[H]
        \centering
        \[\begin{aligned}
            \bracket{\textsf{L}_{\textit{sd}}} ::={}& \textit{sd\textsuperscript{v}} \quad \text{when } \textit{sd\textsuperscript{v}} \leq_\mathsf{s} \textit{sd} \\
                            % &\mid \texttt{sbrs}  \\
            \bracket{\textsf{S}_{\textit{sd}}} ::={}& \bracket{\textsf{L}_{\textit{sd}}} \\
                            &\mid \texttt{lit } \bracket{\text{integer}} \\
            \bracket{\textsf{R}_{\textit{sd}}} ::={}& \bracket{\textsf{S}_{\textit{sd}}} \\
                            &\mid \bracket{ \text{unary operator} } \bracket{ \textsf{S}_{\textit{sd}}}\\
                            &\mid \bracket{\textsf{S}_{\textit{sd}}} \bracket{\textsf{binary operator}} \bracket{ \textsf{S}_{\textit{sd}} }
        \end{aligned}\]
        \caption{Target language grammar (part 1)}
        \label{fig: target_grammar_1}
    \end{figure}

    \begin{figure}[H]
        \centering
        \[\begin{aligned}
            \bracket{\textsf{I}_{\textit{sd}}} ::={}& \texttt{stop} \\
                            &\left.
                            \begin{aligned}
                            &|{\ } \bracket{\textsf{L}_{\textit{sd}+\delta}} := \bracket{\textsf{R}_{\textit{sd}}}[\delta] \text{ ; } \bracket{\textsf{I}_{\textit{sd}+\delta}} \\
                            % &|{\ } \texttt{if } \bracket{\textsf{S}_{\textit{sd}}} \bracket{\text{relational operator}} \bracket{\textsf{S}_{\textit{sd}}}[\delta] \\
                            % & \hphantom{\mid {}} \texttt{ then } \bracket{\textsf{I}_{\textit{sd}+\delta}} \texttt{ else } \bracket{\textsf{I}_{\textit{sd}+\delta}} \\
                            &|{\ } \texttt{adjustdisp} [\delta] \text{ ; } \bracket{\textsf{I}_{\textit{sd}+\delta}} \\
                            \end{aligned}
                            \quad \right\} \text{if } d + \delta \geq 0  \\
                            &{}\mid{} \texttt{popto } \textit{sd′} \text{ ; } \bracket{\textsf{I}_{\textit{sd′}}} \quad \text{when } \textit{sd′} \leq_\mathsf{s} \textit{sd} \\
        \end{aligned}\]
        \caption{Target language grammar (part 2)}
        \label{fig: target_grammar_2}
    \end{figure}
    Some explanations of the grammar are as follows:
    \begin{itemize}
        % \item 
        %     $\texttt{sbrs}$: a register used to communicate the result of function procedures implemented by closed subroutines.
        \item 
            $\texttt{lit}$: a literal integer.
        \item 
            $\texttt{stop}$: stops the execution of the program.
        \item 
            $\delta$: an integer representing the displacement adjustment.
        \item 
            $\texttt{adjustdisp}$: adjusts the displacement of specified stack descriptor.
        \item 
            $\texttt{popto}$: pops to the specified stack descriptor, which is used to reduce the number of frames.
    \end{itemize}
    Note that the design of right-hand sides ensures that the right operand of the assignment contains at most one operator, which means that any instruction that involves multiple operators must be divided into multiple instructions. This feature of the target language significantly complicates the compilation of integer expressions, which we will analysis in detail in~\secref{subsubsec: compiler_intexp}.

@@

    \section{Customised Library} \label{sec: lib}
    The customised library is a collection of types and functions centred around natural numbers, their operations and properties. Instead of using the standard library~\autocite{agda_std}, I implemented a customised library for the following reasons:
    \begin{itemize}
        \item
            \textbf{Specialised Requirements}: The standard library's definitions (e.g. subtraction as monus, `$\dotdiv$') do not align with my project's need for a rigorously defined subtraction operation. This is specified in~\secref{subsec: subtraction}.
        \item
            \textbf{Verification Clarity}: By defining the basic data types and functions from scratch, I can tailor properties (e.g. the property `$n-[n-m] \equiv m$') to my specific requirements, and avoid unnecessary complexity (e.g. `$\leq$' being also defined as a boolean predicate in the standard library). 
        \item
            \textbf{Minimal Dependencies}: Avoiding the standard library reduces external assumptions, making the project self-contained. 
    \end{itemize}
    All definitions and properties that are listed in this section can be fount in Appendix~\ref{app: lib_properties}.

    \subsection{Definitions from standard library}
    The following is a list of definitions (or their variants) from the standard library and PLFA~\cite{plfa} that are redefined in the customised library:
    \begin{itemize}
        \item 
            Natural numbers ($\mb{\bN}$), integers ($\mb{\bZ}$) and addition (\mbt{+}).
        \item
            Equality (\mbt{≡}), congruence ($\mb{\mathsf{cong}}$), substitution ($\mb{\mathsf{sub}}$) and transitivity of equality ($\mb{\mathsf{trans}}$).
        \item
            Order of natural numbers, including \mbt{≤} and \mbt{<}.
        \item
            Properties of preorder and addition (summarised in Table~\ref{tab: properties}).
    \end{itemize}
    \begin{table}[H]
        \centering
        \begin{tabular}{|l|l|}
            \hline
            \textbf{Agda term} & \textbf{Mathematical meaning} \\
            \hhline{|=|=|}
            \mbt{+-identity\textsuperscript{r}} & $\forall n \in \bN.\ n + 0 = n$ \\
            \hline
            \mbt{+-suc\textsuperscript{r}}& $\forall m, n \in \bN.\ m + \textsf{suc } n = \textsf{suc } (m + n)$ \\
            \hline
            \mbt{+-comm\textsuperscript{r}} & $\forall m, n \in \bN.\ m + n \equiv n + m$ \\
            \hline
            \mbt{≤-irrelevant} & For all $m, n \in \bN$, we consider all proofs of $m \leq n$ to be equal. \\
            \hline
            \mbt{≤-refl} & $\forall n \in \bN.\ n \leq n$ \\
            \hline
            \mbt{≤-trans} & $\forall m, n, p \in \bN.\ m \leq n \land n \leq p \Rightarrow m \leq p$ \\
            \hline
            \mbt{n≤suc-n} & $\forall n \in \bN.\ n \leq \textsf{suc } n$ \\
            \hline
            \mbt{m≡n,p≤n}\mb{$\to$}\mbt{p≤m} & $\forall p, m, n \in \bN.\ m \equiv n \land p \leq n \Rightarrow p \leq m$ \\
            \hline
            \mbt{+}\mb{$\to$}\mbt{≤} & $\forall m, n \in \bN.\ m \leq m + n$ \\
            \hline
            \mbt{+}\mb{$\to$}\mb{≤\textsuperscript{r}} & $\forall m, n \in \bN.\ m \leq n + m$ \\
            \hline
            \mbt{<-trans} & $\forall m, n, p \in \bN.\ m < n \land n < p \Rightarrow m < p$ \\
            \hline
            \mbt{<}\mb{$\to$}\mbt{s≤} & $\forall m, n \in \bN.\ m < n \Rightarrow \textsf{suc } m \leq n$ \\
            \hline
            \mbt{<}\mb{$\to$}\mbt{≤} & $\forall m, n \in \bN.\ m < n \Rightarrow m \leq n$ \\
            \hline
            \mbt{≤-compare} & $\forall m, n \in \bN.\ m \leq n \lor n \leq m$ \\
            \hline
        \end{tabular}
        \caption{Properties of order and addition}
        \label{tab: properties}
    \end{table}
    The properties are defined in a minimalistic way and tailored to the requirements of the compiler. For example, \mbt{≤-compare} in Table~\ref{tab: properties} is a property that states that for all $m, n \in \bN$, either $m \leq n$ or $n \leq m$, which is used to case split in the implementation of the compiler. In the standard library, a similar property is defined to split the cases into $m < n$, $m \equiv n$ and $m > n$, which adds unnecessary complexity to the implementation, as all the proof required for compiler is in the form of $m \leq n$.
    % \subsection{Natural numbers, integers and addition} \label{subsec: natural_numbers}
    % The definition of natural numbers, integers and addition is standard, and it is defined as follows:
    % \[\includegraphics[scale=1.2]{/lib/natural.pdf}\]
    % The use of \og{\textsf{BUILTIN}} is specified in~\secref{subsubsec: builtin_pragma}.

@@

        Please see Appendix~\ref{app: compiler_details} for the full implementation of the function \textsf{NewVar}.

        \subsubsection{Integer expression} \label{subsubsec: compiler_intexp}
        The denotational semantics of type $\mathsf{intexp}$ is an exponential from an integer continuation to a command continuation. We need to fill the integer continuation with appropriate $R$ terms to get the command continuation. 
        \begin{itemize}
            \item For integer literals, we can simply fill in the integer literal (as an $\mathsf{R}$ term) to the integer continuation $\beta$.
            \item For addition and negation, we wish to fill the negation or the sum of the given expression(s) to the integer continuation $\beta$. However, this is only possible when the given expression(s) are simple left-hand sides, as the grammar of the target language in Figure~\ref{fig: target_grammar_1} restricts right-hand sides to contain at most one operator. An auxiliary function \textsf{use-temp} that stores the intermediate result of given expressions is introduced to solve this problem. 
        \end{itemize}
        With \textsf{use-temp}, the denotational semantics of addition and negation is simply wrapping $\beta$ with a lambda function using \textsf{use-temp} to get temporary variable(s) of the given expression(s), and directly fill the negation or the sum of the variable(s) to the integer continuation $\beta$. Note that in the denotational semantics of addition, we need to use the functorial mapping of simple right-hand sides for the first argument, as the stack may grow during the evaluation of the second argument. The denotational semantics of integer expressions is defined as follows:
        \[\includegraphics[scale=1.2]{/compiler/term_intexp.pdf}\]
        The auxiliary function \textsf{use-temp} checks if the given expression is a simple left-hand side,
        \begin{itemize}
            \item If it is, we simply fill the expression to the integer continuation $\beta$.\footnote{The type of $\beta$ here is actually a bit different to integer continuation I defined, as it only accepts simple right-hand sides instead of right-hand sides. Reynolds gave a wrong definition of the type of this term. I have corrected it in my Agda implementation.}
            \item If it is not, we need to use a temporary variable to store the result of the expression, and then fill the temporary variable to the integer continuation $\beta$.
        \end{itemize}
        However, regarding the stack descriptor, according to Reynolds' model, we assume
        \begin{itemize}
            \item \textit{sd} is the stack descriptor before the evaluation of plus or negation, which is the position of the temporary variable in the stack.
            \item \textit{sd′} is the stack descriptor after the evaluation on the given expression and just before the assignment of the temporary variable. \textit{sd} ≤ₛ \textit{sd′} is guaranteed.
            \item \textit{sd″} = \textit{sd} +ₛ 1 is the stack descriptor after the assignment of the temporary variable, as it is just large enough to hold the temporary variable.
        \end{itemize}
        In Reynolds' model, the stack descriptor can be directly adjusted from \textit{sd′} to \textit{sd″}. However, in the Agda implementation, we need to know whether the adjustment is positive or negative and use \textsf{assign-inc} or \textsf{assign-dec} respectively. This is done by using \mbt{≤-compare} in Table~\ref{tab: properties}. We also need to show that the adjustment is valid for each case. Please refer to Appendix~\ref{app: compiler_details} for the full implementation of the function \textsf{use-temp}.
        % The auxiliary function \textsf{use-temp} is defined as follows:
        % \[\includegraphics[scale=1.2]{/compiler/assign_use-temp.pdf}\]
